---
title: "P8157_final project"
author: "Hongzhu Ren"
date: "2023-12-10"
output: html_document
---

```{r}
library(tidyverse)
library(haven)
library(table1)
library(nlme)
library(geepack)
library(IrregLong)
library(survival)
library(gtsummary)
```


```{r}
cd4 <- read_dta("./data/cd4.dta")|>
  rename(treatment = group)
```

## EDA anlaysis

### Table1
```{r}
follow_up <- cd4|>
  group_by(id)|>
  summarise(
    follow_up = n()
  )

cd4_tidy <- cd4|>
  mutate(
    treatment = factor(treatment,
                       levels=c(1:4),
                       labels = c("alternating 400mg didanosine monthly",
                                  "+ 2.25mg zalcitabine", 
                                  "+ 400mg didanosine", 
                                  "+ 400mg didanosine & 400mg nevirapine")),
    sex = case_match(
      sex,
      1 ~ "Male",
      0 ~ "Female"
    )
  )|>
  rename(gender = sex)|>
  left_join(follow_up)

cd4_bl <- cd4_tidy|>
  filter(week==0)
```


```{r}
## identify subjects without baseline measurement
id_bl <- cd4_bl|>
  pull(id)|>
  unique()

id <- cd4|>pull(id)|>unique()

weird <- cd4|>filter(id %in% setdiff(id,id_bl))
```


```{r}
label(cd4_bl$treatment) <- "Treatment"
label(cd4_bl$age) <- "Age"
label(cd4_bl$gender) <- "Gender"
label(cd4_bl$follow_up) <- "Follow-up times"
label(cd4_bl$logcd4) <- "Log(CD4 counts+1)"

caption  <- "Baseline stats"
footnote <- "Observations within 40 weeks"

table1(~age+gender+follow_up+logcd4 | treatment, data = cd4_bl,caption = caption, footnote = footnote)
```

* subjects without baseline measurement
* skewed in gender

## spaghetti plot

#### Treatment and gender
```{r}
cd4_tidy |>
  mutate(
    id = factor(id)
  )|>
  ggplot(aes(x = week, y = logcd4, color = id))+
  geom_line()+
  geom_smooth(color = "red")+
  facet_grid(cols = vars(treatment),rows = vars(gender))+
  theme(legend.position = "none")

```

From group 1 to 4, the overall decline in log(cd4 count+1) decrease. 

Group 4 of males even have an increase trend in log(cd4 count+1)


#### Treatment and age
```{r}
cd4_tidy |>
  mutate(
    id = factor(id),
    age = case_when(age>=10&age<30 ~ "10~30",
                    age>=30&age<50 ~ "30~50",
                    age>=50 ~ ">50"
                    ),
    age = factor(age,level = c("10~30","30~50",">50"))
  )|>
  ggplot(aes(x = week, y = logcd4, color = id))+
  geom_line()+
  geom_smooth(color = "red")+
  facet_grid(cols = vars(treatment),rows = vars(age))+
  theme(legend.position = "none")
```

* Lower age has lower reduce rate.
* For all treatment group, there is average wave trend in age 10-30.
* In general, cd4 count decrease eventually for treatment1 and treatment2.
* Effect of treatment 3 varied in different age group.
* Treatment4 has an average control of decrease of cd4 for all age group


Little difference in average log(cd4+1) in each treatment group


```{r}
hist(cd4_tidy$age)
```


```{r}
cd4_reg <- cd4|>
  mutate(
    treatment = factor(treatment,
                       levels=c(1:4),
                       labels = c("1",
                                  "2", 
                                  "3", 
                                  "4")),
    sex = case_match(
      sex,
      1 ~ "Male",
      0 ~ "Female"
    ),
    age = case_when(age>=10&age<20 ~ "10",
                    age>=20&age<30 ~ "20",
                    age>=30&age<40 ~ "30",
                    age>=40&age<50 ~ "40",
                    age>=50&age<60 ~ "50",
                    age>=60 ~ "60"),
    age = as.numeric(age)
  )|>
  rename(gender = sex)
```

```{r back up}
age = case_when(age>=10&age<20 ~ "10",
                    age>=20&age<30 ~ "20",
                    age>=30&age<40 ~ "30",
                    age>=40&age<50 ~ "40",
                    age>=50&age<60 ~ "50",
                    age>=60&age<70 ~ "60",
                    age>=70 ~ "70")
```




## LMM model

#### naive model with random intercept
```{r}
LMM.M.fit1 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~1|id),
                data = cd4_reg,
                method = "ML")

summary(LMM.M.fit1)
```


#### selected model with random intercept
```{r}
LMM.M.fit.s <- lme(logcd4~treatment*week+age+gender-treatment,
                random = reStruct(~1|id),
                data = cd4_reg,
                method = "ML")

summary(LMM.M.fit.s)
```


#### selected model with independent random slope and random intercept
```{r}
LMM.M.fit2 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~treatment|id, pdClass = "pdDiag"),
                data = cd4_reg,
                method = "ML")
 
summary(LMM.M.fit2)
```

```{r random slope& intercept}
coe <- test$coefficients$random$id

coe <- as.data.frame(coe)

coe_plot <- coe |>
  mutate(
    treatment2 = ifelse(treatment2==0,NA,treatment2),
    treatment3 = ifelse(treatment3==0,NA,treatment3),
    treatment4 = ifelse(treatment4==0,NA,treatment4)
  )

coe_plot|>
  filter(!is.na(treatment4))|>
  ggplot(aes(y = treatment4, x = `(Intercept)`))+
  geom_point()
```

* Difference in standard deviation of the random effect of each treatment group, suggest heteroskedasticity

#### selected model with correlated random slope and random intercept
```{r}
LMM.M.fit3 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~treatment|id),
                data = cd4_reg,
                method = "ML")

summary(LMM.M.fit3)
```


#### selected model random intercept + heteroskedasticity across treatment
```{r}
LMM.M.fit4 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~1|id),
                weights = varIdent(form = ~1| treatment),
                data = cd4_reg,
                method = "ML")

test <- summary(LMM.M.fit4)
test$tTable|>
  round(4)|>
  knitr::kable()
```


#### selected model random intercept + heteroskedasticity across age
```{r}
LMM.M.fit5 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~1|id),
                weights = varIdent(form = ~1| age),
                data = cd4_reg,
                method = "ML")

summary(LMM.M.fit5)
```


#### selected model random intercept + heteroskedasticity across gender
```{r}
LMM.M.fit6 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~1|id),
                weights = varIdent(form = ~1| gender),
                data = cd4_reg,
                method = "ML")

summary(LMM.M.fit6)
```


#### selected model random intercept and auto-related
```{r}
LMM.M.fit7 <- lme(logcd4~treatment*week+age+gender*week,
                random = reStruct(~1|id),
                correlation = corAR1(form = ~week|id),
                data = cd4_reg,
                method = "ML")

summary(LMM.M.fit7)
```


#### compare models

```{r}
LMM.result <- tibble(
  name = c("LMM.M.fit1","LMM.M.fit2","LMM.M.fit3","LMM.M.fit4","LMM.M.fit5","LMM.M.fit6","LMM.M.fit7"),
  model = list(LMM.M.fit1,LMM.M.fit2,LMM.M.fit3,LMM.M.fit4,LMM.M.fit5,LMM.M.fit6,LMM.M.fit7),
  AIC = map(model, AIC)
)

LMM.result|>
  select(-model)|>
  unnest(name,AIC)|>
  knitr::kable()


```


#### Checking the best fit

```{r}
cd4_reg |>
  mutate(
    epsHat <- resid(LMM.M.fit1,type = "normalized")
  ) |>
  ggplot(aes(x = treatment, y = epsHat))+
  geom_boxplot()

LRT <- -2*(logLik(LMM.M.fit1)-logLik(LMM.M.fit2))

gammaHat <- ranef(LMM.M.fit2)[,1]
```



```{r}
simulate.res <- simulate.lme(LMM.M.fit1,
                             nsim = 1000,
                             seed = 0,
                             m2 = LMM.M.fit2)

# compute LRT statistics quantile
LRT_test <- -2*(logLik(LMM.M.fit1)-logLik(LMM.M.fit2))|>as.numeric()

# compute 95 percentile for simulated mix chi-square distribution
LRT = -2*(simulate.res[["null"]][["REML"]]-simulate.res[["alt"]][["REML"]])
LRT <- data.frame(LRT)
quantile_95 <- quantile(LRT$logLik,0.95)

## ???how to make this plot???
ggplot(data = LRT,aes(x = logLik) )+
  geom_density()+
  xlim(0,25)+
  geom_vline(xintercept = quantile_95, linetype = "dashed") +
  annotate("text", x = quantile_95, y = 0, label = "95% Quantile", vjust = 1)+
  labs(
    y = "density",
    x = "LRT statistics",
    title = "simulated chi-square distribution"
  )

```

## GEE


```{r}
GEE.fit1 <- geeglm(logcd4~treatment*week+age++gender*week,
                   id = id,
                   data = cd4_reg,
                   corstr = "independence"
                   )

summary(GEE.fit1)
```


```{r}
GEE.fit2 <- geeglm(logcd4~treatment*week+age++gender*week,
                   id = id,
                   data = cd4_reg,
                   corstr = "exchangeable"
                   )

summary(GEE.fit2)
```


```{r}
GEE.fit3 <- geeglm(logcd4~treatment*week+age+gender*week,
                   id = id,
                   data = cd4_reg,
                   corstr = "ar1"
                   )

summary(GEE.fit3)
```


## Irregular observations

```{r}
cd4_iiw <- cd4_reg|>
  mutate(
    event = 1-as.numeric(is.na(logcd4))
  )|>
  lagfn(c("week","logcd4"), "id",lagfirst = c(0,0))

```

```{r}
hist(cd4$logcd4)
```

#### Calculate inverse intensity weight

```{r}
i <- iiw.weights(Surv(week.lag,week,event)~ gender +
                   logcd4.lag +
                   cluster(id),id="id",time="week",event="event",data=cd4_iiw,
                 invariant=c("id","age","gender","treatment"),lagvars=c("week","logcd4"),maxfu=40,
                 lagfirst=c(0,0),first=FALSE)

tbl_regression(i$m)
```

#### fit iiw GEE

```{r}
iiwgee <- iiwgee(logcd4 ~ treatment*week+age+gender*week,
                 Surv(week.lag,week,event)~gender + logcd4.lag + cluster(id),
                 formulanull=NULL,id="id",time="week",event="event",data=cd4_iiw,
                 invariant=c("id","age","gender","treatment"),lagvars=c("week","logcd4"),maxfu=40,
                 lagfirst=c(0,0),first=FALSE)

summary(iiwgee$geefit)
```

